---
title: "Assignment 8: Time Series Analysis"
author: "Hope Liu"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r setup2, message=FALSE, warning=FALSE}
# Checking packages
library(tidyverse)
library(dplyr)
library(agricolae)
library(lubridate)
library(ggplot2)
library(ggridges)
library(viridis)
library(RColorBrewer)
library(colormap)
library(ggthemes)
library(formatR)
library(trend)
library(zoo)
library(Kendall)
library(tseries)
library(here)
here()
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
options(scipen = 4)

# Generating my theme
Hope.theme <- theme_base()  + 
  theme(
    line = element_line(color = 'darkgreen', linewidth = 2),
    rect = element_rect(color = 'darkgreen'),
    text = element_text(size = 12, color = 'darkgreen', face = "italic"),
    
    plot.title = element_text(size = 14, color = 'darkgreen', face = "bold", margin=margin(b=8)),
    axis.text = element_text(size = 12, color = 'darkgreen', face = "italic"),
    
    axis.ticks = element_line(color = 'darkgreen'),
    panel.grid.major =  element_line(color='cornsilk3', size=0.5),
    panel.grid.minor =  element_blank(),
    
    plot.background = element_rect(fill = 'azure2'),
    panel.background =  element_rect(fill = 'azure2'),
    legend.background = element_rect(fill = 'azure2'),
    legend.key = element_rect(fill = 'azure2'),
    legend.title = element_text(size = 12, face = "bold"),
    legend.position = 'none',
    
    complete = TRUE)

# Set my theme as default theme
theme_set(Hope.theme)
```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#1 Reading files
NC2010 <- read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv"), stringsAsFactors = TRUE)
NC2011 <- read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv"), stringsAsFactors = TRUE)
NC2012 <- read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv"), stringsAsFactors = TRUE)
NC2013 <- read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv"), stringsAsFactors = TRUE)
NC2014 <- read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv"), stringsAsFactors = TRUE)
NC2015 <- read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv"), stringsAsFactors = TRUE)
NC2016 <- read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv"), stringsAsFactors = TRUE)
NC2017 <- read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv"), stringsAsFactors = TRUE)
NC2018 <- read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv"), stringsAsFactors = TRUE)
NC2019 <- read.csv(here("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv"), stringsAsFactors = TRUE)

#Combining to a single dataframe
GraingerOzone <- bind_rows(NC2010, NC2011, NC2012, NC2013, NC2014, NC2015, NC2016, NC2017, NC2018, NC2019)
```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3 Set date to date format
GraingerOzone$Date <- as.Date(GraingerOzone$Date, format = "%m/%d/%Y")
# Check date class
class(GraingerOzone$Date)

# 4 Wrangling
GraingerOzone <- GraingerOzone %>% 
  select(Date, Daily.Max.8.hour.Ozone.Concentration, DAILY_AQI_VALUE)

# 5 Generating a data sequence, make that into a dataframe
Days <- as.data.frame(
  seq(as.Date("2010-01-01"), as.Date("2019-12-31"), by = "day"))
# Changing column name
colnames(Days) <- "Date"

# 6 Joining GraingerOzone to Days
GraingerOzone <- left_join(Days, GraingerOzone, by = 'Date')
```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r, warning=FALSE, message=FALSE, fig.height=6, fig.width=10}
#7 Making plot
plot1 <- GraingerOzone %>% 
ggplot(aes(x=Date, y=Daily.Max.8.hour.Ozone.Concentration)) +
  geom_line() +
  geom_smooth(method = "lm")+
  xlab("Years") +
  ylab("Ozone Concentration in ppm")+
  ggtitle("2010-2019 Ozone concentrations at Grainger High School")
plot1
```

>Answer: **From the plot, it can be seen that ozone level fluctuates year by year. From a decade look, there is a very slight decreasing trend.**


## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8 Interpolating NAs
GraingerOzone <- GraingerOzone %>% 
  mutate(Ozone.Clean = zoo::na.approx(Daily.Max.8.hour.Ozone.Concentration)) %>% 
  mutate(AQI.Clean = zoo::na.approx(DAILY_AQI_VALUE))
```

> Answer: 

**We are not using the piecewise constant because this method fills the missing data with exact the same value as the nearest neighbor, but we do want to our filling of data fit the trend of the broader range of data, which means we want our filled data be different from neighbors.**

**We are not using the spline interpolation because this method could give a data that too fluctuated, making the data series change beyond the day-to-day trend.**

**Using linear interpolation is a simple and straightforward way to provide smooth fillings for the time-series data set.**



9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r, message=FALSE}
#9 Calculating mean Ozone concentration for each year's each month
GraingerOzone.monthly <- GraingerOzone %>% 
  mutate(Month = month(Date)) %>% 
  mutate(Year = year(Date)) %>% 
  group_by(Month, Year) %>% 
  summarise(Monthly.Mean.Ozone = mean(Ozone.Clean))

# Mutating new date column
GraingerOzone.monthly <- GraingerOzone.monthly %>% 
  mutate(Date = my(paste0(Month,"-",Year)))
```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10 Generating two time series objects
GraingerOzone.daily.ts <- ts(GraingerOzone$Ozone.Clean, frequency = 365)

GraingerOzone.monthly.ts <- ts(GraingerOzone.monthly$Monthly.Mean.Ozone, start = c(2010, 1), frequency = 12)
```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r, fig.height=5, fig.width=8}
#11 Decomposing
Daily.Decompose <- stl(GraingerOzone.daily.ts, s.window = "periodic")
plot(Daily.Decompose)
Monthly.Decompose <- stl(GraingerOzone.monthly.ts, s.window = "periodic")
plot(Monthly.Decompose)
```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12 Run SMK test
Monthly.Trend <- Kendall::SeasonalMannKendall(GraingerOzone.monthly.ts)

# Inspect results
Monthly.Trend
summary(Monthly.Trend)
```

> Answer: **In this case, the seasonal Mann-Kendall is the most appropriate, because first, our data trend, as the decomposition plots show, is not linear, so the linear regression would not be suitable. From the seasonal plot we can see somewhat variations between observations in accordance to seasonal changes. So a SMK test to take account of seasonality would be the most appropriate.**


13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r, message=FALSE, fig.height=6, fig.width=10}
# 13 Plotting
plot2 <-
ggplot(GraingerOzone.monthly, aes(x = Date, y = Monthly.Mean.Ozone)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = lm)+
  xlab("Time") +
  ylab("Ozone Concentration in ppm")+
  ggtitle("2010-2019 Monthly Ozone concentrations at Grainger High School")
plot2
```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: **From the statistical result of the SMK time series analysis, the p-value is 0.163, which is > 0.05, which means we would not reject the null hypothesis of this study that there is no significant trend or seasonal pattern on ozone concentrations at the Grainger High school. From the graph we can also see a variation pattern that support the alternative hypothesis. From the smoothed line, there is a slightly decreasing trend, corresponded with the -0.1 tau score reported by the time series model. Within each year, there is clear seasonal pattern that ozone level in cold months are lower than that in hot months.**



15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r, warning=FALSE, fig.height=6, fig.width=10}
#15 Extracting seasonal component, and make it into a dataframe
Seasonal.Component <- as.data.frame(Monthly.Decompose$time.series[,1:3])

Seasonal.Component <- mutate(Seasonal.Component,
        Observed = GraingerOzone.monthly$Monthly.Mean.Ozone,     
        Date = GraingerOzone.monthly$Date)

# Visualize how the seasonal cycle maps onto the data
plot3 <-
ggplot(Seasonal.Component) +
  geom_line(aes(y = Observed, x = Date),  linewidth = 0.5) +
  geom_line(aes(y = seasonal, x = Date), color = "red") +
  geom_hline(yintercept = 0, lty = 2) +
  xlab("Time") +
  ylab("Ozone Concentration in ppm")+
  ggtitle("2010-2019 Monthly Ozone concentrations at Grainger High School, with the Seasonal Component")
plot3

#16 Run MK test
Monthly.Trend.MK <- Kendall::MannKendall(GraingerOzone.monthly.ts)

# Inspect results
Monthly.Trend.MK
summary(Monthly.Trend.MK)
```

> Answer: 

**The non-seasonal MK time series analysis returned the p-value 0.088, which is smaller than the p-value of the SMK analysis we ran in previous step (pSMK = 0.163). Though this 0.088 p-value is still > 0.05, the no-trend null hypothesis would not be rejected. This MK analysis result could support the general slightly decreasing trend of ozone level from 2010 to 2019 in the Grainger High School (tau = -0.105).**

**In short, MK result could only reflect overall data trend through out ten years, SMK result reflects both broad trend and seasonal patterns.**
