---
title: "Assignment 10: Data Scraping"
author: "Hope Liu"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE, warning =FALSE}
#1 Checking packages
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggridges)
library(viridis)
library(RColorBrewer)
library(colormap)
library(ggthemes)
library(formatR)
library(rvest)
library(dataRetrieval)
library(tidycensus)
library(here); here()
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
options(scipen = 4)

# Generating my theme
Hope.theme <- theme_base()  + 
  theme(
    line = element_line(color = 'darkgreen', linewidth = 2),
    rect = element_rect(color = 'darkgreen'),
    text = element_text(size = 12, color = 'darkgreen', face = "italic"),
    
    plot.title = element_text(size = 14, color = 'darkgreen', face = "bold", margin=margin(b=8)),
    axis.text = element_text(size = 12, color = 'darkgreen', face = "italic"),
    
    axis.ticks = element_line(color = 'darkgreen'),
    panel.grid.major =  element_line(color='cornsilk3', size=0.5),
    panel.grid.minor =  element_blank(),
    
    plot.background = element_rect(fill = 'azure2'),
    panel.background =  element_rect(fill = 'azure2'),
    legend.background = element_rect(fill = 'azure2'),
    legend.key = element_rect(fill = 'azure2'),
    legend.title = element_text(size = 12, face = "bold"),
    legend.position = 'none',
    
    complete = TRUE)

# Set my theme as default theme
theme_set(Hope.theme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
Webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
Webpage
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
Water.System.Name <- Webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
Water.System.Name

PWSID <- Webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID

Ownership <- Webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
Ownership

MGD <- Webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
MGD
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data, fig.height=7, fig.width=10, message=FALSE}
#4 Create a dataframe
Durham.WLSP.df <- data.frame("Month" = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12),
                             "Year" = rep(2022,12),
                             "Monthly_Withdrawl_MGD" = as.numeric(MGD))

Durham.WLSP.df <- Durham.WLSP.df %>% 
  mutate(Water_System_Name = !!Water.System.Name,
         PWSID = !!PWSID,
         Ownership = !!Ownership,
         Date = my(paste(Month,"-",Year)))

#5 
ggplot(Durham.WLSP.df,aes(x=Date,y=Monthly_Withdrawl_MGD)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2022 Monthly Maximum Daily Water Use (MGD) for",Water.System.Name),
       y="Withdrawal (mgd)",
       x="Date")
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
scrape.it <- function(the_year, the_PWSID) {
  
  the_website <- read_html(paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", the_PWSID, "&year=", the_year))
  
  the_system_name <- the_website %>% 
    html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
    html_text()
  the_PWSID <- the_website %>% 
    html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
    html_text()
  the_ownership <- the_website %>% 
    html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
    html_text()
  the_MGD <- the_website %>% 
    html_nodes("th~ td+ td") %>% 
    html_text()

df <- data.frame(
  "Month" = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12),
  "Year" = rep(the_year,12),
  "Monthly_Withdrawl_MGD" = as.numeric(the_MGD)) %>%
  mutate(Water_System_Name = !!the_system_name,
         PWSID = !!the_PWSID,
         Ownership = !!the_ownership,
         Date = my(paste(Month,"-",Year)))
return(df)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data, fig.height=7, fig.width=10, message=FALSE}
#7 Run the function
Durham.2015.WLSP.df <- scrape.it(2015, '03-32-010')
view(Durham.2015.WLSP.df)

#Plot
ggplot(Durham.2015.WLSP.df,aes(x=Date,y=Monthly_Withdrawl_MGD)) + 
   geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2015 Monthly Maximum Daily Water Use (MGD) for Durham"),
       y="Withdrawal (mgd)",
       x="Date")
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data, fig.height=7, fig.width=10, message=FALSE}
#8 Run the function
Asheville.2015.WLSP.df <- scrape.it(2015, '01-11-010')
view(Asheville.2015.WLSP.df)

#Plot
ggplot(Asheville.2015.WLSP.df,aes(x=Date,y=Monthly_Withdrawl_MGD)) + 
   geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2015 Monthly Maximum Daily Water Use (MGD) for Asheville"),
       y="Withdrawal (mgd)",
       x="Date")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data, fig.height=6, fig.width=10, message=FALSE}
#9 
the_years = rep(2010:2021)
thee_PWSID = '01-11-010'

#Use purrr's map function
Ash.dfs <- map(the_years,scrape.it,the_PWSID=thee_PWSID)

#Conflate the returned dataframes into a single dataframe
Ash.df <- bind_rows(Ash.dfs)

#Plot
ggplot(Ash.df,aes(x=Date,y=Monthly_Withdrawl_MGD)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2010 - 2021 Monthly Maximum Daily Water Use (MGD) for Asheville"),
       y="Withdrawal (mgd)",
       x="Date")
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

> Answer: **By observing the smooth line, Asheville's water usage shows a slightly decreasing trend from 2010 to 2015, after that, the water usage obviously increased and rising to a higher step than previous years around 2020.**
